Fractional Types

### 2016-12-15 typespp.tex
working on sec 4

\[\begin{array}{c}
\Rule{}
{}
{\jdg{}{}{\idiso : \tau \iso \tau}}
{}
\qquad\qquad
\Rule{}
{\jdg{}{}{c_1 : \tau_1 \iso \tau_2} \quad \vdash c_2 : \tau_2 \iso \tau_3}
{\jdg{}{}{c_1 \odot c_2 : \tau_1 \iso \tau_3}}
{}
\\
\\
\Rule{}
{\jdg{}{}{c_1 : \tau_1 \iso \tau_2} \quad \vdash c_2 : \tau_3 \iso \tau_4}
{\jdg{}{}{c_1 \oplus c_2 : \tau_1 \oplus \tau_3 \iso \tau_2 \oplus \tau_4}}
{}
\qquad\qquad
\Rule{}
{\jdg{}{}{c_1 : \tau_1 \iso \tau_2} \quad \vdash c_2 : \tau_3 \iso \tau_4}
{\jdg{}{}{c_1 \otimes c_2 : \tau_1 \otimes \tau_3 \iso \tau_2 \otimes \tau_4}}
{}
\end{array}\]

@@

The $\Pi$ family of languages is based on type isomorphisms. In the
variant we consider, the set of types $\tau$ includes the empty
type~$\zt$, the unit type $\ot$, and sum $\oplus$ and product
$\otimes$ types. The values classified by these types are the
conventional ones: $\unitv$ of type~$\ot$, $\inl{v}$ and $\inr{v}$ for
injections into sum types, and $(v_1,v_2)$ for product types. The
language has two other syntactic categories of programs to be
described in detail:
\[\begin{array}{lrcl}
(\textrm{Types}) &
  \tau &::=& \zt \alt \ot \alt \tau_1 \oplus \tau_2 \alt \tau_1 \otimes \tau_2 \\
(\textrm{Values}) &
  v &::=& \unitv \alt \inl{v} \alt \inr{v} \alt (v_1,v_2) \\
(\textrm{1-combinators}) &
  c &:& \tau_1 \iso \tau_2 ~ [\textit{see Fig.~\ref{pi-combinators}}] \\
(\textrm{2-combinators}) &
  \alpha &:& c_1 \isotwo c_2 \mbox{~where~} c_1, c_2 : \tau_1 \iso \tau_2
  ~[\textit{see Fig.~\ref{pi-combinators2}}]
\end{array}\]

@@

***WAVEFRONT***

We are now ready to turn the groupoid constructions from the previous section
into an actual programming language. The language $\Pi^/$ will be an extension
of $\Pi$ with new type constructors and combinators for creating and
manipulating syntactic counterparts to $\order{p}$ and $\iorder{p}$.

We begin by adding four new syntactic categories to the definition of $\Pi$ in
Sec.~\ref{opsempi}:

\[\begin{array}{lrcl}
(\textrm{Types}) & \tau &::=& \ldots \\
(\textrm{Values}) & v &::=& \ldots \\
(\textrm{1-combinators}) & c &:& \tau_1 \iso \tau_2 ~::= \ldots \\
(\textrm{2-combinators}) & \alpha &:& c_1 \isotwo c_2 \mbox{~where~}
  c_1, c_2 : \tau_1 \iso \tau_2 ~::= \ldots \\
\\
(\textrm{Fractional Types}) & \mathbb{T} &::=&
  \order{c} \alt \iorder{c} \alt \mathbb{T}_1 \circledast \mathbb{T}_2 \\
  \order{c} \alt \iorder{c} \alt \mathbb{T}_1 \circledast \mathbb{T}_2 \\
 (\textrm{Fractional Values}) & \mathbb{V} &::=&
  c^k \alt \alpha^k \alt \langle \mathbb{V},\mathbb{V} \rangle \\
 (\textrm{/-combinators}) & \rho &:& \mathbb{T}_1 \leftrightspoon \mathbb{T}_2 ~::= \\
& &&     \idiso/ \alt \rho \circledcirc \rho \alt \rho \circledast \rho \\
& &\alt& \unitetl/ \alt \unititl/ \alt \unitetr/ \alt \unititr/ \\
& &\alt& \swapt/ \alt \assoclt/ \alt \assocrt/ \\
& &\alt& \order{\alpha} \alt \eta- \alt \eta+ \alt \epsilon+ \alt \epsilon- \\
(\textrm{//-combinators}) & \varrho &:& \rho_1 \leftrightblackspoon \rho_2 \mbox{~where~}
  \rho_1, \rho_2 : \mathbb{T}_1 \leftrightspoon \mathbb{T}_2 ~::= \ldots
\end{array}\]

>>>>

\[\begin{array}{c}
\Rule{}
{}
{\jdg{}{}{\idiso : \tau \iso \tau}}
{}
\qquad\qquad
\Rule{}
{\jdg{}{}{c_1 : \tau_1 \iso \tau_2} \quad c_2 : \tau_2 \iso \tau_3}
{\jdg{}{}{c_1 \odot c_2 : \tau_1 \iso \tau_3}}
{}
\\
\\
\Rule{}
{\jdg{}{}{c_1 : \tau_1 \iso \tau_2} \quad c_2 : \tau_3 \iso \tau_4}
{\jdg{}{}{c_1 \oplus c_2 : \tau_1 \oplus \tau_3 \iso \tau_2 \oplus \tau_4}}
{}
\qquad\qquad
\Rule{}
{\jdg{}{}{c_1 : \tau_1 \iso \tau_2} \quad c_2 : \tau_3 \iso \tau_4}
{\jdg{}{}{c_1 \otimes c_2 : \tau_1 \otimes \tau_3 \iso \tau_2 \otimes \tau_4}}
{}
\end{array}\]

@@

The $\Pi$ family of languages is based on type isomorphisms. In the
variant we consider, the set of types $\tau$ includes the empty
type~$\zt$, the unit type $\ot$, and sum $\oplus$ and product
$\otimes$ types. The values classified by these types are the
conventional ones: $\unitv$ of type~$\ot$, $\inl{v}$ and $\inr{v}$ for
injections into sum types, and $(v_1,v_2)$ for product types. The
language has two other syntactic categories of programs to be
described in detail.

\begin{definition}[$\Pi$]
\label{def:pi}
The syntax of $\Pi$ is given by the following categories:
\[\begin{array}{lrcl}
(\textrm{Types}) &
  \tau &::=& \zt \alt \ot \alt \tau_1 \oplus \tau_2 \alt \tau_1 \otimes \tau_2 \\
(\textrm{Values}) &
  v &::=& \unitv \alt \inl{v} \alt \inr{v} \alt (v_1,v_2) \\
(\textrm{1-combinators}) &
  c &:& \tau_1 \iso \tau_2 ~ [\textit{see Fig.~\ref{pi-combinators}}] \\
(\textrm{2-combinators}) &
  \alpha &:& c_1 \isotwo c_2 \mbox{~where~} c_1, c_2 : \tau_1 \iso \tau_2
  ~[\textit{see Fig.~\ref{pi-combinators2}}]
\end{array}\]
\end{definition}

@@

We are now ready to turn the groupoid constructions from the previous section into an extension of $\Pi$ with new type constructors and combinators for creating and manipulating syntactic counterparts to $\order{p}$ and $\iorder{p}$.

\begin{figure}[ht]
\[\begin{array}{rrcll}
\unitetl/ :&  \order{\idiso} \circledast \mathbb{T} & \leftrightspoon & \mathbb{T} &: \unititl/ \\
\unitetr/ :&  \mathbb{T} \circledast \order{\idiso} & \leftrightspoon & \mathbb{T} &: \unititr/ \\
\swapt/ :&  \mathbb{T}_1 \circledast \mathbb{T}_2 & \leftrightspoon & \mathbb{T}_2 \circledast \mathbb{T}_1 &: \swapt/ \\
\assoclt/ :&  \mathbb{T}_1 \circledast (\mathbb{T}_2 \circledast \mathbb{T}_3) & \leftrightspoon & (\mathbb{T}_1 \circledast \mathbb{T}_2) \circledast \mathbb{T}_3
  &: \assocrt/ \\
\\
\eta- :& \order{\idiso} & \leftrightspoon & \iorder{c} \circledast \order{c} &: \epsilon- \\
\eta+ :& \order{\idiso} & \leftrightspoon & \order{c} \circledast \iorder{c} &: \epsilon+ \\
\end{array}\]

\[\begin{array}{c}
\Rule{}
{}
{\jdg{}{}{\idiso/ : \mathbb{T} \leftrightspoon \mathbb{T}}}
{}
\qquad
\Rule{}
{\jdg{}{}{\rho_1 : \mathbb{T}_1 \leftrightspoon \mathbb{T}_2} \quad \rho_2 : \mathbb{T}_2 \leftrightspoon \mathbb{T}_3}
{\jdg{}{}{\rho_1 \circledcirc \rho_2 : \mathbb{T}_1 \leftrightspoon \mathbb{T}_3}}
{}
\\
\\
\Rule{}
{\jdg{}{}{\alpha : c_1 \isotwo c_2}}
{\jdg{}{}{\order{\alpha} : \order{c_1}  \leftrightspoon \order{c_2}}}
{}
\qquad
\Rule{}
{\jdg{}{}{\rho_1 : \mathbb{T}_1 \leftrightspoon \mathbb{T}_2} \quad \rho_2 : \mathbb{T}_3 \leftrightspoon \mathbb{T}_4}
{\jdg{}{}{\rho_1 \circledast \rho_2 : \mathbb{T}_1 \circledast \mathbb{T}_3 \leftrightspoon \mathbb{T}_2 \circledast \mathbb{T}_4}}
{}
\end{array}\]

Each combinator $\rho$ has an inverse.
\caption{$\Pi^/$ fractional combinators}
\label{pifrac:comb}
\end{figure}

We call the new extension $\Pi^/$. We begin by adding four new syntactic categories to the definition of $\Pi$ in Sec.~\ref{opsempi}:

\begin{definition}[$\Pi^/$]
\label{def:pifrac}
The syntax of $\Pi^/$ is given by the following categories:
\[\begin{array}{lrcl}
(\textrm{Types}) & \tau &::=& [\textit{see Def.~\ref{def:pi}}] \\
(\textrm{Values}) & v &::=& [\textit{see Def.~\ref{def:pi}}] \\
(\textrm{1-combinators}) & c &:& \tau_1 \iso \tau_2 ~::= [\textit{see Def.~\ref{def:pi}}] \\
(\textrm{2-combinators}) & \alpha &:& c_1 \isotwo c_2 \mbox{~where~}
  c_1, c_2 : \tau_1 \iso \tau_2 ~::= [\textit{see Def.~\ref{def:pi}}] \\
\\
(\textrm{Fractional Types}) & \mathbb{T} &::=&
  \order{c} \alt \iorder{c} \alt \mathbb{T}_1 \circledast \mathbb{T}_2 \\
 (\textrm{Fractional Values}) & \mathbb{V} &::=&
  c^k \alt 1/c^k \alt \langle \mathbb{V},\mathbb{V} \rangle \\
 (\textrm{/-combinators}) & \rho &:& \mathbb{T}_1 \leftrightspoon \mathbb{T}_2 ~::=
  [\textit{see Fig.~\ref{pifrac:comb}}] \\
(\textrm{//-combinators}) & \varrho &:& \rho_1 \leftrightblackspoon \rho_2 \mbox{~where~}
  \rho_1, \rho_2 : \mathbb{T}_1 \leftrightspoon \mathbb{T}_2 ~::=
  [\textit{see Fig.~\ref{pifrac:twocomb}}]
\end{array}\]
\end{definition}
